The imprint of global magma oceans on exoplanet demographics

### 2023-03-07 src/tex/ms.tex
test additions
Add two paragraphs on stellar characterization.

    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using Bioverse, a simulation framework that leverages contextual information from the overall planet population.
    $\ldots$

@@

% Main body
\section{Introduction}
...just dropping material and references for now...
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
Bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and Nautilus are \\
Possibility of follow-up to measureme masses
''}

@@

\section{Methods}\label{sec:methods}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
\todo[inline]{Start methods description with brief statement of goal and overview of the steps. The first sentence is similar, but would be good to separate goal from method here.
}
We determined -- for different survey configurations -- the confidence level with which the runaway greenhouse threshold can be detected in simulated exoplanet observations.
For this purpose we expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected the runaway greenhouse signal, simulated observations of this sample, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).
\todo[inline]{Bioverse will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with Bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:mo_model}.
        The second block simulates an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, the third block deals with testing the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis given the assumed survey design.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}

@@

\subsubsection{Stellar sample from Gaia DR3}
%Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
\todo[inline]{KEVIN: describe stellar population}
...

@@

\subsection{Survey simulations}
The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
   This task includes selection of the targets, application of detection biases, and conducting simulated measurements.
   All of the above are specific to the particular survey.

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/corner.pdf}
        \caption{
            Retrieved posterior distribution of key parameters in the optimistic scenario. The density maps in each panel show relationships between and marginalized distributions of the model parameters as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets. True values of the parameters for the injected effect are shown in orange. The threshold instellation can be reasonably constrained; the predominant water fraction and the fraction of planets with runaway greenhouse climates are degenerate.
        }
        \label{fig:cornerplot}
    \end{centering}
\end{figure}
With such a strong signal, we can attempt an inference of the parameters defining the injected effect.
Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
The threshold instellation can be accurately constrained with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated.

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_statpwr_H2O-f.pdf}
        \caption{
            Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
            Higher values in either parameter result in a more reliable detection.
            For water mass fractions $\gtrsim 10^{-4}$, the statistical power largely depends on the fraction of greenhouse climate planets in the sample.
        }
        \label{fig:statpwr_H2O-f}
    \end{centering}
\end{figure}
Figure~\ref{fig:statpwr_H2O-f} explores the statistical power achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
It is highest for large water inventories and large dilution factors.
 For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
As long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.

\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}\label{sec:res_samplesize}
%\todo{Introduce survey simulation(s). PLATO, Cheops, (Ariel, LIFE/LIFE precursor, Nautilus)\\ What are threshold missions that are able to detect this signal? =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
%\subsubsection{PLATO}
\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
\plato's definition study report~\citep{plato2017} further states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
We base our assessment of \plato's diagnostic power on these estimates.

    \begin{figure*}[ht!]
%    \script{figurescript.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/plato_fgrid.pdf}
            \caption{
            Expected delta-evidences as a function of the fraction of planets having runaway greenhouse climates for different versions of the \plato\ survey.
            The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
                Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
               Even smaller samples are unlikely to yield a significant detection.
            \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the runaway greenhouse signal.
            }
            \label{fig:plato_fgrid}
        \end{centering}
    \end{figure*}

@@

\subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and measured their capability to recover the trend and constrain its parameters.
We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.

@@

\subsubsection{Spectral type of target stars}\label{sec:results_FGK_M}
Since receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
%($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
The samples are volume and magnitude-limited to reflect the target counts of \plato's provisional Long-duration Observation Phase fields~\citep[$15996$ FGK stars in the P1 and P2 samples, $33948$ M~stars in the P4 sample, ][]{Nascimbeni2022}.
The resulting M~dwarf planet sample is significantly larger with $\var{N_M}\pm \var{N_M_err}$ planets compared to $\var{N_FGK}\pm \var{N_FGK_err}$ planets in the FGK sample.
No significant detection is possible in the pure FGK sample, independent of the assumed geophysical parameters.
In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.2$, similar to the case above where all spectral types are considered.

@@

Accurate retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$.
If, on the other hand, a follow-up campaign adds planetary mass estimates and we fit the signal in bulk density space, useful constraints are possible under more pessimistic conditions.
A sample containing only planets around M~dwarf yields a comparable performance as the overall sample.

@@

\subsection{Statistical imprint of exoplanet climates}
\begin{note}
The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
Using Bioverse to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.

@@

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular beyond it, i.e., in the optimistic habitable zone.
The transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.

@@

\subsubsection{Water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent weight fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contants can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

@@

Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
Using Bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
    \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements. For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend. Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements. This can reduce the required planet yield by about a factor of three.  An optimized survey will further focus on M~dwarfs to ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.

@@

\begin{large}\textit{Author contributions:}\end{large}
%M.S., D.A., and T.L. conceived the project.
%M.S. wrote the manuscript and carried out the hypothesis tests and statistical analyses.
%...
%K.H.-U. provided the star generator based on the Gaia catalog.
%G.B. implemented the planet generator in the Bioverse framework.
%All authors provided comments and suggestions on the manuscript.

@@

\software{
Bioverse~\citep{Bixel2021},
Astropy~\citep[][]{AstropyCollaboration2018},
NumPy~\citep[][]{Harris2020},
SciPy~\citep[][]{Virtanen2020},
corner.py~\citep{Foreman-Mackey2016b},
dynesty~\citep{Speagle2020}.
}

>>>>

    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
    $\ldots$

@@

% Main body
\section{Introduction}
...just dropping material and references for now...
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and Nautilus are \\
Possibility of follow-up to measureme masses
''}

@@

\section{Methods}\label{sec:methods}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
\todo[inline]{Start methods description with brief statement of goal and overview of the steps. The first sentence is similar, but would be good to separate goal from method here.
}
We determined -- for different survey configurations -- the confidence level with which the runaway greenhouse threshold can be detected in simulated exoplanet observations.
For this purpose we expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected the runaway greenhouse signal, simulated observations of this sample, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).
\todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:mo_model}.
        The second block simulates an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, the third block deals with testing the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis given the assumed survey design.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}

@@

\subsubsection{Stellar sample from Gaia DR3}
%Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
%\todo[inline]{KEVIN: describe stellar population}
\todo[inline]{KEVIN: update reference to my paper once accepted}
The original \texttt{Bioverse} stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function. Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible for Hardegree-Ullman et al. (submitted) to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \texttt{Bioverse}. Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived for \texttt{Bioverse}, but for a full description, please refer to Hardegree-Ullman et al. (submitted).

Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{GCNS2021} as the basis for deriving stellar parameters for the \texttt{Bioverse} catalog. The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands \citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS \citep{Cutri2003}, all of which were used in stellar classification. From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars. Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}. Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction. Stellar radii were then computed with the effective temperatures and lumonisities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs. Finally, masses were derived from either the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim0.7\,M_{\odot}$, and that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs. The derived stellar parameters from Hardegree-Ullman et al. (submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.

@@

\subsection{Survey simulations}
The survey module of \bioverse\ converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
This task includes selection of the targets, application of detection biases, and conducting simulated measurements, all of which are specific to the particular survey.
For each planet-level measurement such as radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/corner.pdf}
        \caption{
            Retrieved posterior distribution of key parameters in the optimistic scenario. The density maps in each panel show relationships between and marginalized distributions of the threshold instellation $S_\mathrm{thresh}$, the water mass fraction $x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$ as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets. True values of the parameters for the injected effect are shown in orange. The threshold instellation can be reasonably constrained; the predominant water fraction and the fraction of planets with runaway greenhouse climates are degenerate.
        }
        \label{fig:cornerplot}
    \end{centering}
\end{figure}
With such a strong signal, we can attempt an inference of the parameters defining the injected effect.
Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
The threshold instellation can be accurately constrained with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated.

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_statpwr_H2O-f.pdf}
        \caption{
            Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
            Higher values in either parameter result in a more reliable detection.
            For water mass fractions $\gtrsim 10^{-4}$, the statistical power largely depends on the fraction of greenhouse climate planets in the sample.
        }
        \label{fig:statpwr_H2O-f}
    \end{centering}
\end{figure}
Figure~\ref{fig:statpwr_H2O-f} explores the statistical power of the hypothesis test achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
It is highest for large water inventories and large dilution factors.
 For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
As long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.

\subsection{Detecting the runaway greenhouse transition with \plato}\label{sec:res_samplesize}
%\todo{Introduce survey simulation(s). PLATO, Cheops, (Ariel, LIFE/LIFE precursor, Nautilus)\\ What are threshold missions that are able to detect this signal? =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
%\subsubsection{PLATO}
\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
\plato's definition study report~\citep{plato2017} further states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
We base our assessment of \plato's diagnostic power on these estimates.

    \begin{figure*}[ht!]
%    \script{figurescript.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/plato_fgrid.pdf}
            \caption{
            Expected delta-evidences as a function of the fraction of planets with runaway greenhouse climates for different versions of the \plato\ survey.
            The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
                Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
               Even smaller samples are unlikely to yield a significant detection.
            \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the runaway greenhouse signal.
            }
            \label{fig:plato_fgrid}
        \end{centering}
    \end{figure*}

@@

\subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.

@@

\subsubsection{Spectral type of target stars}\label{sec:results_FGK_M}
Since the receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
%($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
The samples are volume and magnitude-limited to reflect the target counts of \plato's provisional Long-duration Observation Phase fields~\citep[$15996$ FGK stars in the P1 and P2 samples, $33948$ M~stars in the P4 sample, ][]{Nascimbeni2022}.
The resulting M~dwarf planet sample is significantly larger with $\var{N_M}\pm \var{N_M_err}$ planets compared to $\var{N_FGK}\pm \var{N_FGK_err}$ planets in the FGK sample.
No significant detection is possible in the pure FGK sample, independent of the assumed geophysical parameters.
In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.2$, similar to the case above where all spectral types are considered.

@@

Accurate retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$.
If, on the other hand, a follow-up campaign adds planetary mass estimates and we fit the signal in bulk density space, useful constraints are possible under more pessimistic conditions.
A sample containing only planets around M~dwarf yields a performance comparable to that of the overall sample.

@@

\subsection{Statistical imprint of exoplanet climates}
\begin{note}
The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.

@@

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.

@@

\subsubsection{Water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent weight fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

@@

Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
    \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements. For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend. Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements. This can reduce the required planet yield by about a factor of three.  An optimized survey will further focus on M~dwarfs to ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.

@@

\begin{large}\textit{Author contributions:}\end{large}
%M.S., D.A., and T.L. conceived the project.
%M.S. wrote the manuscript and carried out the hypothesis tests and statistical analyses.
%...
%K.H.-U. provided the star generator based on the Gaia catalog.
%G.B. implemented the planet generator in the \bioverse\ framework.
%All authors provided comments and suggestions on the manuscript.

@@

\software{
\bioverse ~\citep{Bixel2021},
Astropy~\citep[][]{AstropyCollaboration2018},
NumPy~\citep[][]{Harris2020},
SciPy~\citep[][]{Virtanen2020},
corner.py~\citep{Foreman-Mackey2016b},
dynesty~\citep{Speagle2020}.
}
